Crime in Austin Texas: Exploration, Classification, and Trends

### 2024-06-14 code/processing-code/processingfile.qmd
Significant amount more cleaning done. Created a processed data file and saved to data folder. Removed several template files.

---
title: "An example cleaning script"
author: "Andreas Handel"
date: "2024-02-07"
output: html_document
editor: 
  markdown: 
    wrap: sentence
---

# Processing script

### Setup

Load needed packages.\
*Pacman::p_load* makes sure they are installed before auto-loading.\
*tidyverse* for general data processing/cleaning.\
*lubridate* for cleaning/re-coding date fields.\
*skimr* for nice visualization of data.\
*here* to set paths.

```{r}
#install.packages(pacman)
pacman::p_load(tidyverse,lubridate,skimr,here)
```

# Data loading

```{r}
# path to data
# note the use of the here() package and not absolute paths
data_location <- here::here("data","raw-data","raw_crime_060424.csv")
rawdata <- read.csv(data_location, TRUE)
```

# Check data

First we can look at the codebook, which can be found [here](https://data.austintexas.gov/Public-Safety/Crime-Reports/fdj4-gpfu/about_data).
I will recreate it here as it currently exists, and I will have a copy of the pdf attachment codebook included in the repo as well.
However, please note if this analysis is ever repeated that careful consideration should be taken to ensure field have not been changed, removed, or redefined.

| Field                       | Description                                                                                                  |
|------------------------------|------------------------------------------|
| Incident Number             | Incident report number                                                                                       |
| Highest Offense Description | Description                                                                                                  |
| Highest Offense Code        | Code                                                                                                         |
| Family Violence             | Incident involves family violence? Y = yes, N = no                                                           |
| Occurred Date Time          | Date and time (combined) incident occurred                                                                   |
| Occurred Date               | Date the incident occurred                                                                                   |
| Occurred Time               | Time the incident occurred                                                                                   |
| Report Date Time            | Date and time (combined) incident was reported                                                               |
| Report Date                 | Date the incident was reported                                                                               |
| Report Time                 | Time the incident was reported                                                                               |
| Location Type               | General description of the premise where the incident occurred                                               |
| Address                     | Incident location                                                                                            |
| Zip code                    | Zip code where incident occurred                                                                             |
| Council District            | Austin city council district where the incident occurred                                                     |
| APD Sector                  | APD sector where incident occurred                                                                           |
| APD District                | APD district where incident occurred                                                                         |
| PRA                         | APD police reporting area where incident occurred                                                            |
| Census Tract                | Census tract where incident occurred                                                                         |
| Clearance Status            | How/whether crime was solved (see Clearance lookup)                                                          |
| Clearance Date              | Date crime was solved                                                                                        |
| UCR Category                | Code for the most serious crimes identified by the FBI as part of its Uniform Crime Reporting program        |
| Category Description        | Description for the most serious crimes identified by the FBI as part of its Uniform Crime Reporting program |
| X-coordinate                | X-coordinate where the incident occurred                                                                     |
| Y-coordinate                | Y-coordinate where incident occurred                                                                         |
| Latitude                    | Latitude where incident occurred                                                                             |
| Longitude                   | Longitude where the incident occurred                                                                        |
| Location                    | 3rd party generated spatial column                                                                           |

| Clearance lookup |                      |
|------------------|----------------------|
| C                | Cleared by Arrest    |
| O                | Cleared by Exception |
| N                | Not cleared          |

Several ways of looking at the data

```{r}
dplyr::glimpse(rawdata)
summary(rawdata)
head(rawdata)
skimr::skim(rawdata)
unique(rawdata$Location.Type)
```

# Cleaning

By inspecting the data as done above, we find some problems that need addressing:

```{r}
d1 <- rawdata %>% dplyr::filter( Height != "sixty" ) %>% 
                  dplyr::mutate(Height = as.numeric(Height))
skimr::skim(d1)
hist(d1$Height)
```

Now we see that there is one person with a height of 6.
That could be a typo, or someone mistakenly entered their height in feet.
Since we unfortunately don't know, we might need to remove this person, which we'll do here.

```{r}
d2 <- d1 %>% dplyr::mutate( Height = replace(Height, Height=="6",round(6*30.48,0)) )
skimr::skim(d2)
```

Height values seem ok now.

Now let's look at the `Weight` variable.
There is a person with weight of 7000, which is impossible, and one person with missing weight.
To be able to analyze the data, we'll remove those individuals as well.

```{r}
d3 <- d2 %>%  dplyr::filter(Weight != 7000) %>% tidyr::drop_na()
skimr::skim(d3)
```

Now checking the `Gender` variable.
Gender should be a categorical/factor variable but is loaded as character.
We can fix that with simple base R code to mix things up.

```{r}
d3$Gender <- as.factor(d3$Gender)  
skimr::skim(d3)
```

Now we see that there is another NA, but it's not `NA` from R, instead it was loaded as character and is now considered as a category.
Well proceed here by removing that individual with that NA entry.
Since this keeps an empty category for Gender, I'm also using droplevels() to get rid of it.

```{r}
d4 <- d3 %>% dplyr::filter( !(Gender %in% c("NA","N")) ) %>% droplevels()
skimr::skim(d4)
```

All done, data is clean now.

Let's assign at the end to some final variable, this makes it easier to add further cleaning steps above.

```{r}
processeddata <- d4
```

# Save data

Finally, we save the clean data as RDS file.
I suggest you save your processed and cleaned data as RDS or RDA/Rdata files.
This preserves coding like factors, characters, numeric, etc.
If you save as CSV, that information would get lost.
However, CSV is better for sharing with others since it's plain text.
If you do CSV, you might want to write down somewhere what each variable is.

See here for some suggestions on how to store your processed data: http://www.sthda.com/english/wiki/saving-data-into-r-data-format-rds-and-rdata

```{r}
save_data_location <- here::here("data","processed-data","processeddata.rds")
saveRDS(processeddata, file = save_data_location)
```

# Notes

Removing anyone who had "faulty" or missing data is one approach.
It's often not the best.
based on your question and your analysis approach, you might want to do cleaning differently (e.g. keep individuals with some missing information).

>>>>

version https://git-lfs.github.com/spec/v1
oid sha256:8ffd2c5fe0b3ccdc979dec853978e42ac882fefa8612df567f1f65d7abde7670
size 14120

### 2024-06-14 products/manuscript/Manuscript.qmd
created processing file to start data cleaning and exploration. Added codebook for reference. renamed data file to be in line with naming conventions for Repo.
Significant amount more cleaning done. Created a processed data file and saved to data folder. Removed several template files.

---
title: "Crime in Austin Texas: Exploration, Classification, and Trends"
subtitle: ""
author: William Hyltin
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/apa.csl
---

The structure below is one possible setup for a data analysis project (including the course project). For a manuscript, adjust as needed. 
You don't need to have exactly these sections, but the content covering those sections should be addressed.

This uses MS Word as output format. [See here](https://quarto.org/docs/output-formats/ms-word.html) for more information. You can switch to other formats, like html or pdf. See [the Quarto documentation](https://quarto.org/) for other formats.

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
```

# Summary/Abstract
_Write a summary of your project._

{{< pagebreak >}}

# Introduction

## General Background Information
_Provide enough background on your topic that others can understand the why and how of your analysis_ 
[Last year](https://www.fox7austin.com/news/austin-named-city-with-15th-biggest-homicide-rate-problem-study), Austin was ranked among the top US cities with a problem in homicide rate. Earlier [this year](https://www.kvue.com/article/news/local/austin-crime-rate-drops-2024-2020-comparison/269-b62042b7-a6f2-4224-87ff-ede655a37912), the claim was made that crime has dropped to a new low since 2021. The intention of this analysis will be to investigate both claims, and gain a better understanding of Austin's criminal activity.
What the analysis will not cover would be things like *causes* of the changes in crime, at least not directly. I may identify shifts that could inform the cause, but the goal will not be to drill down into every potential socioeconomic factor and pick the primary drivers, instead simply to observe what crime in the city of Austin looks like and what we might be able to expect going forward. Aside from resources outlined in our course resources, I was hoping to also draw from this [book](https://otexts.com/fpp2/index.html). It's not one I have used before but it had been recommended to me previously and I would like to learn more about time series and forecasting in general.

## Description of data and data source
_Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section._
My primary data source will come from [here](https://data.austintexas.gov/Public-Safety/Crime-Reports/fdj4-gpfu/about_data), but I will likely need to pull from other sources as well, potentially finding other cities' crime stats that made the same list of problematic homicide rates, or even Austin's demographic information and statistics like population size.
```{r Load Raw Data}
d1 <- read.csv(here('data','raw-data','Crime_Reports_20240604.csv'))
```

```{r Quick Summaries}
summary(d1)
head(d1)
```
The data several data and time fields, many of which are character strings meaning they will need to be cleaned/recoded. The rest of the fields are mostly classification variables and location information, but with 2,470,867 there is quite a bit of opportunity for frequency and classification analysis.

## Questions/Hypotheses to be addressed
_State the research questions you plan to answer with this analysis._
Ultimately, I would like to explore a few questions: 
  1. Has crime truly dropped, or is it expected to rise again as the year progresses? 
  2. Has the homicide rate dropped with crime? 
  3. Has the homicide rate been a problem for long time, or was it truly an emerging problem last year?

To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above and have the right bibtex key. Then you can include like this:

Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].

{{< pagebreak >}}

# Methods

_Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement._

## Schematic of workflow

Sometimes you might want to show a schematic diagram/figure that was not created with code (if you can do it with code, do it). @fig-schematic is an example of some - completely random/unrelated - schematic that was generated with Biorender.
We store those figures in the `assets` folder.

```{r}
#| label: fig-schematic
#| fig-cap: "A figure that is manually generated and shows some overview/schematic. This has nothing to do with the data, it's just a random one from one of our projects I found and placed here."
#| echo: FALSE
knitr::include_graphics(here("assets","antigen-recognition.png"))
```

## Data aquisition
_As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next._

## Data import and cleaning
_Write code that reads in the file and cleans it so it's ready for analysis. Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts. If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g. as supplement in a paper). All materials, including files that contain code, should be commented well so everyone can follow along._

## Statistical analysis
_Explain anything related to your statistical analyses._

{{< pagebreak >}}

# Results

## Exploratory/Descriptive analysis

_Use a combination of text/tables/figures to explore and describe your data. Show the most important descriptive results here. Additional ones should go in the supplement. Even more can be in the R and Quarto files that are part of your project._

@tbl-summarytable shows a summary of the data.

Note the loading of the data providing a **relative** path using the `../../` notation. (Two dots means a folder up). You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path. You can also use the `here` R package to create paths. See examples of that below. I generally recommend the `here` package.

```{r}
#| label: tbl-summarytable
#| tbl-cap: "Data summary table."
#| echo: FALSE
resulttable=readRDS("../../results/tables/summarytable.rds")
knitr::kable(resulttable)
```

## Basic statistical analysis

_To get some further insight into your data, if reasonable you could compute simple statistics (e.g. simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable. Though note that unless you pre-specified the outcome and main exposure, any "p<0.05 means statistical significance" interpretation is not valid._

@fig-result shows a scatterplot figure produced by one of the R scripts.

```{r}
#| label: fig-result
#| fig-cap: "Height and weight stratified by gender."
#| echo: FALSE
knitr::include_graphics(here("results","figures","height-weight-stratified.png"))
```

## Full analysis

_Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc. This might again be code that is best placed in one or several separate R scripts that need to be well documented. You want the code to produce figures and data ready for display as tables, and save those. Then you load them here._

Example @tbl-resulttable2 shows a summary of a linear model fit.

```{r}
#| label: tbl-resulttable2
#| tbl-cap: "Linear model fit table."
#| echo: FALSE
resulttable2 = readRDS(here("results","tables","resulttable2.rds"))
knitr::kable(resulttable2)
```

{{< pagebreak >}}

# Discussion

## Summary and Interpretation
_Summarize what you did, what you found and what it means._

## Strengths and Limitations
_Discuss what you perceive as strengths and limitations of your analysis._

## Conclusions
_What are the main take-home messages?_

_Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end_

This paper [@leek2015] discusses types of analyses.

These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.

Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header. Many more style files for almost any journal [are available](https://www.zotero.org/styles). You also specify the location of your bibtex reference file in the YAML. You can call your reference file anything you like.

{{< pagebreak >}}

# References

>>>>

version https://git-lfs.github.com/spec/v1
oid sha256:53791af84dddf4d0e7fd1fcad04d4f30f60de0ca33f13bf28575f0f87d12c65f
size 4288
